% ============================================================================
%  AI-Driven Educational Analytics System — Technical Report
%  Intelligent Exam Question Difficulty Prediction using
%  Feature Engineering & Logistic Regression
% ============================================================================

\documentclass[12pt, a4paper]{article}

% --- Packages ---------------------------------------------------------------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{amsmath, amssymb}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{listings}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{tocloft}
\usepackage{parskip}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, calc, fit, backgrounds}

% --- Colour Palette ---------------------------------------------------------
\definecolor{primary}{HTML}{334155}
\definecolor{accent}{HTML}{2563EB}
\definecolor{muted}{HTML}{94A3B8}
\definecolor{codebg}{HTML}{F1F5F9}
\definecolor{linkblue}{HTML}{3B82F6}

% --- Hyperref Setup ---------------------------------------------------------
\hypersetup{
    colorlinks  = true,
    linkcolor   = primary,
    urlcolor    = linkblue,
    citecolor   = accent,
    pdftitle    = {AI-Driven Educational Analytics System — Technical Report},
    pdfauthor   = {Siddharth Dangi},
}

% --- Header / Footer --------------------------------------------------------
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textcolor{muted}{AI-Driven Educational Analytics System}}
\fancyhead[R]{\small\textcolor{muted}{Technical Report}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% --- Section Styling --------------------------------------------------------
\titleformat{\section}
  {\Large\bfseries\color{primary}}{\thesection}{1em}{}[\vspace{-0.5em}\textcolor{muted}{\rule{\textwidth}{0.4pt}}]
\titleformat{\subsection}
  {\large\bfseries\color{primary}}{\thesubsection}{1em}{}
\titleformat{\subsubsection}
  {\normalsize\bfseries\color{primary}}{\thesubsubsection}{1em}{}

% --- Code Listing Style -----------------------------------------------------
\lstdefinestyle{pythonstyle}{
    backgroundcolor=\color{codebg},
    basicstyle=\ttfamily\small,
    keywordstyle=\color{accent}\bfseries,
    commentstyle=\color{muted}\itshape,
    stringstyle=\color{primary},
    breaklines=true,
    frame=single,
    rulecolor=\color{muted},
    numbers=left,
    numberstyle=\tiny\color{muted},
    xleftmargin=2em,
    framexleftmargin=1.5em,
}
\lstset{style=pythonstyle}

% ============================================================================
%  DOCUMENT BEGIN
% ============================================================================
\begin{document}

% --- Title Page -------------------------------------------------------------
\begin{titlepage}
\begin{center}
    \vspace*{2cm}

    {\Huge\bfseries\textcolor{primary}{AI-Driven Educational\\[0.3cm] Analytics System}}\\[0.6cm]
    {\Large\textcolor{muted}{Intelligent Exam Question Difficulty Prediction\\via Feature Engineering \& Logistic Regression}}

    \vspace{1.5cm}

    \textcolor{muted}{\rule{0.6\textwidth}{0.4pt}}

    \vspace{1.5cm}

    {\large
    \textbf{Authors:}\\[0.3cm]
    Siddharth Dangi\\[0.15cm]
    \vspace{0.5cm}
    \textbf{Date:} March 2026\\[0.3cm]
    \textbf{Version:} 1.0
    }

    \vspace{2cm}

    \begin{tikzpicture}
        \node[draw=primary, rounded corners=8pt, line width=1.5pt,
              inner sep=16pt, text width=0.7\textwidth, align=center,
              fill=codebg] {
            \textbf{Abstract}\\[6pt]
            \small
            This report presents an end-to-end machine-learning system that
            automatically predicts the difficulty level of examination questions.
            The system combines TF-IDF text vectorisation with numerical
            student performance features (average score, correct rate, score
            variance) and employs a Logistic Regression classifier within a
            scikit-learn pipeline. The model achieves \textbf{98.9\% accuracy}
            on a 5{,}000-question dataset with balanced class distribution.
            A Streamlit web application is provided for real-time batch and
            single-question difficulty analysis.
        };
    \end{tikzpicture}

    \vfill
    {\small\textcolor{muted}{GenAI Project — Machine Learning Pipeline \& Streamlit Application}}
\end{center}
\end{titlepage}

% --- Table of Contents ------------------------------------------------------
\tableofcontents
\newpage

% ============================================================================
%  1. INTRODUCTION
% ============================================================================
\section{Introduction}\label{sec:intro}

\subsection{Motivation}
Educators, instructional designers, and testing organisations invest considerable manual effort in evaluating the difficulty and quality of examination questions. A misjudged question can skew test results and inaccurately measure student proficiency. Automating this ``first-pass'' quality assurance can accelerate exam development cycles and improve assessment reliability.

\subsection{Objective}
The \textbf{AI-Driven Educational Analytics System} is an end-to-end, production-structured machine-learning pipeline that:
\begin{enumerate}[leftmargin=2em]
    \item Takes raw question text and associated student performance metrics as input.
    \item Applies \textbf{TF-IDF vectorisation} to extract 3{,}000 text features from the question text.
    \item Combines text features with \textbf{3 numerical features} (average score, correct rate, score variance) using a \texttt{ColumnTransformer}.
    \item Classifies each question into a categorical \textbf{difficulty tier}: Easy, Medium, or Hard.
    \item Exposes predictions through an interactive \textbf{Streamlit web application} with rich visualisations.
\end{enumerate}

\subsection{Scope}
\begin{itemize}[leftmargin=2em]
    \item \textbf{Domain:} Multi-subject English-language examination questions spanning Chemistry, Biology, Data Structures, Genetics, and more.
    \item \textbf{Dataset:} 5{,}000 exam questions with associated metadata and student performance statistics.
    \item \textbf{Model:} Logistic Regression within a scikit-learn pipeline.
    \item \textbf{Deployment:} Local Streamlit application with a serialised model artefact (\texttt{model.pkl}).
\end{itemize}

% ============================================================================
%  2. DATASET
% ============================================================================
\section{Dataset}\label{sec:dataset}

\subsection{Source \& Scale}
The dataset comprises \textbf{5{,}000 examination questions} stored in \texttt{data/5000\_dataset\_genai.csv}. Each record contains:
\begin{itemize}[leftmargin=2em]
    \item \texttt{question\_id} — Unique identifier for each question
    \item \texttt{question\_text} — Full text of the examination question
    \item \texttt{topic} — Subject area (e.g., Chemistry, Biology, Data Structures, Genetics)
    \item \texttt{average\_score} — Mean student score on a 0–10 scale
    \item \texttt{correct\_rate} — Proportion of students answering correctly (0–1)
    \item \texttt{score\_variance} — Variance in student scores
    \item \texttt{difficulty\_label} — Ground truth label: Easy, Medium, or Hard
\end{itemize}

\subsection{Dataset Schema}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabularx}{\textwidth}{l l X}
    \toprule
    \textbf{Column} & \textbf{Type} & \textbf{Description} \\
    \midrule
    \texttt{question\_id}       & Integer & Unique question identifier \\
    \texttt{question\_text}     & String  & Full text of the exam question \\
    \texttt{topic}              & String  & Subject/topic category \\
    \texttt{average\_score}     & Float   & Mean student score (0--10) \\
    \texttt{correct\_rate}      & Float   & Correct response proportion (0--1) \\
    \texttt{score\_variance}    & Float   & Variance in student scores \\
    \texttt{difficulty\_label}  & String  & Target variable: Easy / Medium / Hard \\
    \bottomrule
\end{tabularx}
\end{center}

\subsection{Class Distribution}
The dataset exhibits a \textbf{balanced} class distribution, which is advantageous for classifier training:

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lrr}
    \toprule
    \textbf{Class} & \textbf{Count} & \textbf{Proportion} \\
    \midrule
    Easy   & $\sim$1{,}667 & $\sim$33.3\% \\
    Medium & $\sim$1{,}667 & $\sim$33.3\% \\
    Hard   & $\sim$1{,}666 & $\sim$33.3\% \\
    \bottomrule
\end{tabular}
\end{center}

\noindent The balanced distribution eliminates the need for resampling techniques (e.g., SMOTE) or class weighting.

\subsection{Sample Records}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\small
\begin{tabularx}{\textwidth}{l X c c c c}
    \toprule
    \textbf{ID} & \textbf{Question Text} & \textbf{Topic} & \textbf{Avg Score} & \textbf{Correct Rate} & \textbf{Label} \\
    \midrule
    1 & Explain the data normalization in Chemistry\ldots & Chemistry & 8.09 & 0.825 & Easy \\
    2 & Compute the key assumption in Data Structures\ldots & Data Structures & 4.00 & 0.604 & Medium \\
    3 & Analyze the logical equivalence in Biology\ldots & Biology & 1.11 & 0.117 & Hard \\
    \bottomrule
\end{tabularx}
\end{center}

% ============================================================================
%  3. FEATURE ENGINEERING
% ============================================================================
\section{Feature Engineering}\label{sec:features}

The model uses a \texttt{ColumnTransformer} to combine two distinct feature extraction pipelines:

\subsection{Text Features — TF-IDF Vectorisation}
The \texttt{question\_text} column is processed by a \texttt{TfidfVectorizer} with the following configuration:

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
    \toprule
    \textbf{Parameter} & \textbf{Value} \\
    \midrule
    \texttt{max\_features}   & 3{,}000 \\
    Analyser                  & Word-level (default) \\
    Sublinear TF              & False (default) \\
    Norm                      & L2 (default) \\
    Stop words                & None (default) \\
    \bottomrule
\end{tabular}
\end{center}

\noindent TF-IDF (Term Frequency--Inverse Document Frequency) captures the relative importance of words in each question relative to the entire corpus. The resulting feature matrix has dimensions $n \times 3{,}000$.

\subsubsection{TF-IDF Formula}
\begin{equation}
    \text{tfidf}(t, d) = \text{tf}(t, d) \times \log\left(\frac{N}{\text{df}(t)}\right)
\end{equation}

\noindent Where $t$ = term, $d$ = document, $N$ = total documents, $\text{df}(t)$ = number of documents containing $t$.

\subsection{Numerical Features — Standard Scaling}
Three numerical features are normalised using \texttt{StandardScaler}:

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
    \toprule
    \textbf{Feature} & \textbf{Description} \\
    \midrule
    \texttt{average\_score}   & Mean student score (0--10) \\
    \texttt{correct\_rate}    & Proportion of correct responses (0--1) \\
    \texttt{score\_variance}  & Variance in student scores \\
    \bottomrule
\end{tabular}
\end{center}

\subsubsection{Standard Scaling Formula}
\begin{equation}
    z = \frac{x - \mu}{\sigma}
\end{equation}

\noindent Where $\mu$ = feature mean, $\sigma$ = feature standard deviation (computed on training data).

\subsection{Feature Fusion}
The \texttt{ColumnTransformer} horizontally concatenates the TF-IDF matrix ($n \times 3{,}000$) with the scaled numerical matrix ($n \times 3$), producing a final feature matrix of dimensions $n \times 3{,}003$.

\begin{center}
\begin{tikzpicture}[
    node distance=0.8cm and 2.5cm,
    every node/.style={font=\small},
    block/.style={rectangle, draw=primary, rounded corners=4pt, fill=codebg,
                  text width=3.5cm, minimum height=0.9cm, align=center,
                  line width=0.8pt},
    arrow/.style={-{Stealth[length=6pt]}, line width=0.8pt, color=primary}
]

\node[block] (text) {\texttt{question\_text}\\{\scriptsize String column}};
\node[block, below=of text] (nums) {Numerical Columns\\{\scriptsize 3 features}};

\node[block, right=of text] (tfidf) {TF-IDF Vectorizer\\{\scriptsize 3{,}000 features}};
\node[block, right=of nums] (scaler) {Standard Scaler\\{\scriptsize 3 features}};

\node[block, right=3cm of $(tfidf)!0.5!(scaler)$] (concat) {Column\\Transformer\\{\scriptsize 3{,}003 features}};

\node[block, right=of concat] (clf) {Logistic\\Regression};

\draw[arrow] (text) -- (tfidf);
\draw[arrow] (nums) -- (scaler);
\draw[arrow] (tfidf) -- (concat);
\draw[arrow] (scaler) -- (concat);
\draw[arrow] (concat) -- (clf);

\end{tikzpicture}
\end{center}

% ============================================================================
%  4. MODEL ARCHITECTURE
% ============================================================================
\section{Model Architecture}\label{sec:models}

\subsection{Pipeline Overview}
The entire model is encapsulated in a single \textbf{scikit-learn \texttt{Pipeline}} object, ensuring that preprocessing and classification are applied consistently during both training and inference:

\begin{lstlisting}[language=Python, caption={Model Pipeline Definition}]
from sklearn.pipeline import Pipeline
from sklearn.compose import ColumnTransformer
from sklearn.feature_extraction.text import TfidfVectorizer
from sklearn.preprocessing import StandardScaler
from sklearn.linear_model import LogisticRegression

preprocessor = ColumnTransformer(transformers=[
    ("text", TfidfVectorizer(max_features=3000), "question_text"),
    ("num", StandardScaler(),
     ["average_score", "correct_rate", "score_variance"])
])

model = Pipeline([
    ("preprocessing", preprocessor),
    ("classifier", LogisticRegression(max_iter=1000))
])
\end{lstlisting}

\subsection{Classifier --- Logistic Regression}
Logistic Regression was chosen for its:
\begin{itemize}[leftmargin=2em]
    \item \textbf{Interpretability:} Linear decision boundaries are easy to understand.
    \item \textbf{Efficiency:} Fast training even with 3{,}003 features.
    \item \textbf{Probabilistic output:} Produces calibrated class probabilities via the softmax function.
    \item \textbf{Low risk of overfitting:} Implicit L2 regularisation.
\end{itemize}

\subsubsection{Multi-Class Strategy}
The classifier uses the \textbf{One-vs-Rest (OvR)} strategy by default. For each class $c \in \{$Easy, Medium, Hard$\}$, a binary logistic model is trained:

\begin{equation}
    P(y = c \mid \mathbf{x}) = \frac{1}{1 + e^{-(\mathbf{w}_c^\top \mathbf{x} + b_c)}}
\end{equation}

\noindent The predicted class is the one with the highest probability.

\subsection{Hyperparameters}

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{ll}
    \toprule
    \textbf{Parameter} & \textbf{Value} \\
    \midrule
    \texttt{max\_iter}       & 1{,}000 \\
    \texttt{penalty}         & L2 (default) \\
    \texttt{C}               & 1.0 (default) \\
    \texttt{solver}          & lbfgs (default) \\
    \texttt{multi\_class}    & auto (OvR) \\
    \texttt{random\_state}   & Not set (deterministic via data split) \\
    \bottomrule
\end{tabular}
\end{center}

% ============================================================================
%  5. TRAINING & EVALUATION
% ============================================================================
\section{Training \& Evaluation}\label{sec:training}

\subsection{Data Split}
\begin{itemize}[leftmargin=2em]
    \item \textbf{Train:} 80\% of the dataset (4{,}000 samples)
    \item \textbf{Test:} 20\% of the dataset (1{,}000 samples)
    \item \textbf{Stratification:} Stratified split on \texttt{difficulty\_label} to preserve class proportions.
    \item \textbf{Random Seed:} \texttt{random\_state=42} for reproducibility.
\end{itemize}

\subsection{Training Process}
The \texttt{Pipeline.fit()} method processes the data sequentially:
\begin{enumerate}[leftmargin=2em]
    \item \texttt{ColumnTransformer} fits the TF-IDF vocabulary and scaler statistics on training data.
    \item The fused feature matrix is passed to \texttt{LogisticRegression.fit()}.
    \item The trained pipeline is serialised via \texttt{joblib.dump()} as \texttt{model.pkl}.
\end{enumerate}

\subsection{Evaluation Metrics}
\begin{itemize}[leftmargin=2em]
    \item \textbf{Accuracy:} $\frac{\text{correct predictions}}{\text{total predictions}}$
    \item \textbf{Precision} (weighted): $\sum_{c} \frac{|c|}{N} \cdot \frac{\text{TP}_c}{\text{TP}_c + \text{FP}_c}$
    \item \textbf{Recall} (weighted): $\sum_{c} \frac{|c|}{N} \cdot \frac{\text{TP}_c}{\text{TP}_c + \text{FN}_c}$
    \item \textbf{F1-Score} (weighted): Harmonic mean of precision and recall
    \item Per-class precision/recall via \texttt{classification\_report}
    \item Confusion matrix visualisation
    \item ROC curves and Precision-Recall curves (One-vs-Rest)
    \item 5-Fold Cross-Validation accuracy
    \item Learning curve analysis
\end{itemize}

% ============================================================================
%  6. RESULTS
% ============================================================================
\section{Results}\label{sec:results}

\subsection{Overall Classification Performance}

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{l c}
    \toprule
    \textbf{Metric} & \textbf{Score} \\
    \midrule
    Accuracy                & \textbf{98.90\%} \\
    Precision (weighted)    & \textbf{98.90\%} \\
    Recall (weighted)       & \textbf{98.90\%} \\
    F1-Score (weighted)     & \textbf{98.90\%} \\
    \bottomrule
\end{tabular}
\end{center}

\subsection{Per-Class Performance}

\begin{center}
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{l cccc}
    \toprule
    \textbf{Class} & \textbf{Precision} & \textbf{Recall} & \textbf{F1-Score} & \textbf{Support} \\
    \midrule
    Easy   & 0.99 & 1.00 & 0.99 & 334 \\
    Medium & 0.99 & 0.98 & 0.98 & 333 \\
    Hard   & 0.99 & 0.99 & 0.99 & 333 \\
    \midrule
    \textbf{Weighted Avg} & \textbf{0.99} & \textbf{0.99} & \textbf{0.99} & \textbf{1{,}000} \\
    \bottomrule
\end{tabular}
\end{center}

\subsection{Key Observations}

\begin{enumerate}[leftmargin=2em]
    \item \textbf{Near-perfect accuracy across all classes.} The balanced class distribution and strong feature signals (average score, correct rate) contribute to the model's impressive performance.

    \item \textbf{Medium class has the lowest recall (98\%).} This is expected, as Medium questions occupy the boundary region between Easy and Hard, making them slightly harder to classify.

    \item \textbf{TF-IDF adds marginal value over numerical features alone.} The three numerical features (average score, correct rate, score variance) are very strong predictors of difficulty. The TF-IDF features provide additional context from the question text.

    \item \textbf{High model confidence.} The prediction confidence scores are generally above 90\%, indicating that the model is well-calibrated for this dataset.

    \item \textbf{No class imbalance issues.} Unlike many real-world datasets where difficulty classes are skewed, this dataset's balanced distribution results in consistent performance across all classes.
\end{enumerate}

% ============================================================================
%  7. VISUALIZATIONS
% ============================================================================
\section{Model Performance Visualisations}\label{sec:visualizations}

The \texttt{notebook/visualizations.ipynb} notebook generates 14 detailed plots to evaluate the model:

\begin{center}
\renewcommand{\arraystretch}{1.3}
\small
\begin{tabularx}{\textwidth}{c l X}
    \toprule
    \textbf{\#} & \textbf{Visualisation} & \textbf{Purpose} \\
    \midrule
    1  & Overall Metrics Bar Chart & Accuracy, Precision, Recall, F1 at a glance \\
    2  & Confusion Matrix (Counts) & Raw classification results \\
    3  & Confusion Matrix (Normalised) & Percentage-based view \\
    4  & Per-Class Precision/Recall/F1 & Grouped bar comparison \\
    5  & ROC Curves (One-vs-Rest) & AUC per class \\
    6  & Precision-Recall Curves & Average precision per class \\
    7  & Confidence Distribution & Overall confidence histogram \\
    8  & Confidence by Predicted Class & Per-class confidence spread \\
    9  & Correct vs Misclassified & Confidence correlation with correctness \\
    10 & Class Distribution (Train/Test) & Verifies stratified split \\
    11 & Feature Violin Plots & Score, correct rate, variance by difficulty \\
    12 & Learning Curve & Data efficiency analysis \\
    13 & 5-Fold Cross-Validation & Per-fold accuracy consistency \\
    14 & Misclassification Analysis & Error pattern identification \\
    \bottomrule
\end{tabularx}
\end{center}

% ============================================================================
%  8. SYSTEM ARCHITECTURE
% ============================================================================
\section{System Architecture}\label{sec:architecture}

\subsection{High-Level Pipeline}

\begin{center}
\begin{tikzpicture}[
    node distance=1.2cm and 2cm,
    every node/.style={font=\small},
    block/.style={rectangle, draw=primary, rounded corners=4pt, fill=codebg,
                  text width=3.8cm, minimum height=1cm, align=center,
                  line width=0.8pt},
    arrow/.style={-{Stealth[length=6pt]}, line width=0.8pt, color=primary}
]

% Training Pipeline
\node[block] (data) {Raw Data\\{\scriptsize\texttt{5000\_dataset\_genai.csv}}};
\node[block, right=of data] (train) {Model Training\\{\scriptsize\texttt{train.ipynb}}};
\node[block, right=of train] (save) {Serialisation\\{\scriptsize\texttt{model.pkl}}};

% Inference Pipeline
\node[block, below=1.5cm of data] (input) {User Input\\{\scriptsize Streamlit \texttt{app.py}}};
\node[block, right=of input] (pipe) {Pipeline Inference\\{\scriptsize TF-IDF + Scaler + LR}};
\node[block, right=of pipe] (out) {Results Dashboard\\{\scriptsize Metrics + Visualisations}};

% Arrows
\draw[arrow] (data) -- (train);
\draw[arrow] (train) -- (save);
\draw[arrow] (save) -- (pipe);
\draw[arrow] (input) -- (pipe);
\draw[arrow] (pipe) -- (out);

% Labels
\node[above=0.3cm of train, color=muted, font=\footnotesize\bfseries] {TRAINING PIPELINE};
\node[above=0.3cm of pipe, color=muted, font=\footnotesize\bfseries] {INFERENCE PIPELINE};

\end{tikzpicture}
\end{center}

\subsection{Repository Structure}
\begin{lstlisting}[language={}, basicstyle=\ttfamily\footnotesize, frame=none, backgroundcolor=\color{codebg}]
AI-driven_educational_analytics_system/
  README.md                          # Project documentation
  app.py                             # Streamlit web application
  model.pkl                          # Serialised trained model
  confusion_matrix.png               # Generated confusion matrix
  data/
    5000_dataset_genai.csv           # Dataset (5,000 questions)
  notebook/
    train.ipynb                      # Model training notebook
    visualizations.ipynb             # 14 performance visualisations
  report/
    report.tex                       # LaTeX technical report (this file)
\end{lstlisting}

% ============================================================================
%  9. APPLICATION INTERFACE
% ============================================================================
\section{Application Interface}\label{sec:app}

The production application (\texttt{app.py}) is built with \textbf{Streamlit} and provides two tabs:

\subsection{Tab 1: Batch Analytics}
\begin{enumerate}[leftmargin=2em]
    \item The user uploads a CSV file containing exam questions with required columns: \texttt{question\_text}, \texttt{average\_score}, \texttt{correct\_rate}, \texttt{score\_variance}.
    \item The pipeline predicts difficulty labels and confidence scores for all questions.
    \item Seven interactive visualisations are generated:
    \begin{itemize}
        \item Difficulty distribution (pie chart + bar chart)
        \item Confidence score histogram
        \item Feature averages by difficulty (grouped bar chart)
        \item Average score vs.\ correct rate (scatter plot, coloured by difficulty)
        \item Score variance distribution (box plot)
        \item Topic-wise difficulty breakdown (stacked horizontal bar chart)
        \item Confusion matrix (if ground truth column is present)
    \end{itemize}
    \item Results can be downloaded as a CSV file.
\end{enumerate}

\subsection{Tab 2: Single Question Prediction}
\begin{enumerate}[leftmargin=2em]
    \item The user enters question text and three numerical values (average score, correct rate, score variance).
    \item The model returns a predicted difficulty label with a coloured badge.
    \item A confidence progress bar and per-class probability bar chart are displayed.
\end{enumerate}

\subsection{Design}
The interface follows a \textbf{dark-mode} design with custom CSS:
\begin{itemize}[leftmargin=2em]
    \item Background: \texttt{\#0e1117}
    \item Card backgrounds: \texttt{\#1c1f26}
    \item Easy badge: \texttt{\#1f8b4c} (green)
    \item Medium badge: \texttt{\#d4a017} (amber)
    \item Hard badge: \texttt{\#b02a37} (red)
\end{itemize}

% ============================================================================
%  10. LIMITATIONS
% ============================================================================
\section{Limitations \& Known Issues}\label{sec:limitations}

\begin{enumerate}[leftmargin=2em]
    \item \textbf{Post-Exam Dependency:} The model relies on post-exam statistics (average score, correct rate, score variance), which are only available after students have taken the exam. This limits its use for pre-exam difficulty estimation.

    \item \textbf{Text-Only Analysis:} The model cannot parse images, diagrams, or graphs. Questions reliant on visual context may be misclassified.

    \item \textbf{Language Support:} The TF-IDF vectoriser is optimised for English-language questions. Non-English text will produce inaccurate features.

    \item \textbf{Single Model:} Only Logistic Regression is implemented. Ensemble methods (XGBoost, Random Forest) may improve performance on more challenging datasets.

    \item \textbf{Dataset Specificity:} The model is trained on 5{,}000 synthetic questions. Real-world questions with unusual formatting or domain-specific terminology may yield lower-confidence predictions.

    \item \textbf{No Feature Importance Analysis:} Due to the TF-IDF vectorisation producing 3{,}000 sparse features, traditional feature importance interpretation is limited.
\end{enumerate}

% ============================================================================
%  11. FUTURE WORK
% ============================================================================
\section{Future Work}\label{sec:future}

\subsection{Short-Term Improvements}
\begin{itemize}[leftmargin=2em]
    \item Build a \textbf{pre-exam model} using only question text (BERT/sentence transformers) to predict difficulty before administering the exam.
    \item Compare against \textbf{ensemble methods} (XGBoost, Random Forest, SVM) to benchmark performance.
    \item Apply \textbf{SHAP values} for model interpretability and feature importance analysis.
    \item Implement \textbf{hyperparameter tuning} via grid search or Bayesian optimisation.
    \item Add \textbf{confidence thresholding} to flag low-confidence predictions for manual review.
\end{itemize}

\subsection{Long-Term --- Agentic AI Framework}
\begin{itemize}[leftmargin=2em]
    \item \textbf{LLM Reasoning:} Use large language models (Gemini / GPT) to solve questions step-by-step and measure \emph{conceptual} complexity.
    \item \textbf{RAG for Curriculum Alignment:} Retrieval-Augmented Generation to compare questions against educational standards.
    \item \textbf{Multi-Modal Processing:} Vision-Language Models for diagram-dependent questions.
    \item \textbf{Iterative Feedback Loops:} Agentic workflows simulating student failure modes to refine difficulty estimates.
\end{itemize}

% ============================================================================
%  12. CONCLUSION
% ============================================================================
\section{Conclusion}\label{sec:conclusion}

This project demonstrates that a straightforward combination of TF-IDF text vectorisation, numerical feature scaling, and Logistic Regression can achieve \textbf{98.9\% accuracy} in predicting exam question difficulty. The balanced dataset of 5{,}000 questions across Chemistry, Biology, Data Structures, Genetics, and other subjects enables consistent performance across all three difficulty tiers.

The Streamlit application provides an accessible interface for educators to obtain instant difficulty assessments for both individual questions and entire question banks. The pipeline architecture (scikit-learn \texttt{Pipeline} with \texttt{ColumnTransformer}) ensures that preprocessing and inference are fully encapsulated, eliminating feature-engineering leakage and simplifying deployment.

Key next steps include building a pre-exam text-only model, benchmarking against ensemble methods, and integrating SHAP-based interpretability to understand which textual and numerical features drive difficulty predictions.

% ============================================================================
%  APPENDIX
% ============================================================================
\appendix
\section{Technology Stack}\label{app:tech}

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{ll}
    \toprule
    \textbf{Component} & \textbf{Technology} \\
    \midrule
    Language        & Python 3.10+ \\
    ML Framework    & scikit-learn $\geq$ 1.3 \\
    Data Processing & Pandas $\geq$ 2.0, NumPy $\geq$ 1.24 \\
    Visualisation   & Matplotlib $\geq$ 3.7, Seaborn $\geq$ 0.12 \\
    Web Framework   & Streamlit $\geq$ 1.30 \\
    Serialisation   & Joblib $\geq$ 1.3 \\
    \bottomrule
\end{tabular}
\end{center}

\section{Feature Vector Specification}\label{app:features}

The complete feature vector $\mathbf{x} \in \mathbb{R}^{3{,}003}$ used by the deployed model:

\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{c l l c}
    \toprule
    \textbf{\#} & \textbf{Feature} & \textbf{Extractor} & \textbf{Dimensionality} \\
    \midrule
    1--3{,}000 & TF-IDF word features & \texttt{TfidfVectorizer} & 3{,}000 (sparse) \\
    3{,}001 & \texttt{average\_score}   & \texttt{StandardScaler} & 1 \\
    3{,}002 & \texttt{correct\_rate}    & \texttt{StandardScaler} & 1 \\
    3{,}003 & \texttt{score\_variance}  & \texttt{StandardScaler} & 1 \\
    \bottomrule
\end{tabular}
\end{center}

\section{Reproduction Instructions}\label{app:reproduce}

\begin{lstlisting}[language=bash, caption={Full Reproduction Steps}]
# 1. Clone the repository
git clone https://github.com/Siddharth-Dangi/AI-driven_educational_analytics_system.git
cd AI-driven_educational_analytics_system

# 2. Create and activate virtual environment
python3 -m venv venv
source venv/bin/activate

# 3. Install dependencies
pip install streamlit scikit-learn pandas numpy matplotlib seaborn joblib

# 4. Train the model (run all cells in notebook/train.ipynb)
# Or run the training script if available

# 5. Launch the application
streamlit run app.py
# Opens at http://localhost:8501

# 6. Explore model performance
# Open notebook/visualizations.ipynb and run all cells
\end{lstlisting}

\end{document}
